\documentclass{article}
\usepackage[fleqn]{amsmath}
\usepackage{amssymb}
\usepackage{fontspec,xltxtra,xunicode}
\usepackage{fontspec}
\defaultfontfeatures{Scale=MatchLowercase}

%% \setmainfont[Mapping=tex-text]{Times New Roman}
%% \setromanfont[Mapping=tex-text]{Times New Roman}
%% \setsansfont[Mapping=tex-text]{Arial}

\setmainfont[Mapping=tex-text]{TeX Gyre Pagella}
%% \setromanfont[Mapping=tex-text]{TeX Gyre Pagella}
%% \setsansfont[Mapping=tex-text]{TeX Gyre Heros}

\usepackage{tikz}
%% \usepackage{pgfplots}
%% \usepackage{pgfplotstable}
%% %% Preamble:
%% \pgfplotsset{width=7cm,compat=1.9}
%% %% \pgfplotsset{xticklabel={\tick},scaled x ticks=false}
%% %% \pgfplotsset{plot coordinates/math parser=false}

\usetikzlibrary{arrows,shapes,patterns,backgrounds,spy}
\usepackage{pgffor}

%% \usepackage{animate}

%% \usepackage{arrayjobx}
%% \usepackage{multido}

%% \usepackage{layouts}

%% \usepackage{etoolbox}

%% \newcounter{mylistcounter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\large

\section{Exploring the Notation}


\section{Asymmetry of Inductive Definition and Proof}

Standard inductive definition: defines all canonical elements; does
not \emph{define} non-standard elements.  That would not make
intuitive sense.  Why do we think it works for standard elements?  The
reason must be axiomatic - we cannot hope to \emph{prove} directly
that every element of an infinity is in canonical form.  But the
inference from base case and inductive hypothesis to all canonicals is
intuitive and compelling.

Standard inductive proof: proving base and inductive steps suffices as
proof for all canonical elements.  Standard inductive proof is
justified by standard inductive definition.  But it does not extend to
non-canonical elements.  There is no intuitive basis for that idea.

Non-standard inductive definition:  doesn't make sense.

Non-standard inductive proof: a standard inductive proof on canonical
elements suffices as proof for all non-canonical elements.  This must
involve an additional axiom, just like standard inductive definition.
But it doesn't seem very intuitive.  So if non-standard proof makes
sense, why not non-standard definition?

\section{Truth Considered Harmful}

In type theory we do not prove that propositions are true; we prove propositions.

Proof that a proposition is true makes truth look like a property; it
would amount to proof that T(P).  We don't do that.

The proven proposition is the last step in a proof.  That's what it
means to prove a proposition.  We take that to mean the proposition is true.

This perspective is essential if we are to make sense of the fact that
in TT we can prove propositions without proving they are true.  What
we prove is that P is a proposition.

Two disciplines, two perspectives.  On the one hand, we have
mathematics and logic; on the other, the classical and the
anti-classical perspectives.  Under the classical perspective,
denotation is the semantic primitive, and logic is truth-conditional.
Under anti-classical perspectives, inference is the semantic
primitive, and logic is expressive.  I use the umbrella term
``anti-classical'' because a variety of more-or-less distinct
approaches - intuitionism, constructivism, etc. - count as
anti-classical: what they all have in common is rejection (with
varying degrees of strength) of denotational semantics and
truth-conditional logic (among other things).

HoTT is presented as a new foundations of mathematics; it can with
equal justice be considered a new foundations of logic.  The HoTT book
takes a primarily mathematical perspective; in this it follows
Martin-Löff, whose motivation in inventing ITT was to find an
intuitionistic foundation for mathematics.  That explains why his
papers sometimes invoke classical reasoning, for example when he
glosses \(a:A\) as meaning that \(a\) denotes an element of \(A\).  It
also explains why he compares the ``:'' operator to the set membership
operator \(\in\).  He could have compared it instead with the
inference operator \(\vdash\).  The two perspectives are not
incompatible, they just represent mathematical and logical
perspectives on type calculii, as we might expect given Curry-Howard.

In traditional logic, we assume truth (and assertion).  An inference
schema like \&-into is taken to mean ``if A true and B true, then A\&B
true''.  But with TT we must split proposition and truth; we should
thus say ``if (A prop and A true) and (B prop and B true), then (A\&B
prop and A\&B true).''  But we use sequents instead of isolated
propositions: ``if a:A and b:B then (a,b):AxB''.  So such rules
involve moving from sequent to sequent.  This allows us to express the
idea that a construction can be proven to be a proposition without
saying anything about its truth; we just write ``P:U'' or ``A->U'' or
the like.  This is something we cannot do in classic FOL. (?)

That bears repeating: TT allows us to say things that we cannot even
express in FOL.  Which means it is inherently higher-order.  (Which
means the HoTT book's account of logic needs revision; its account of
FOL depends on 2nd order logic, since it involves working with
propositions as objects.)

ML's distinction  between proposition and judgment  does not withstand
critical scrutiny.  It depends on  an untenable notion of the relation
between propositional content and  illocutionary force.  The latter is
something  we do,  not something  that  is an  essential component  of
propositions.

ML's schematic diagram of ``A is true'': A is a proposition, A is true
is a judgment.  Nonsense; both are propositions.  Also silly:
obviously ``A is true'' can only be a judgment if it is judge to be
true: ``'A is true' is true''.  Hello, infinite regress.

ML: the premises and conclusion of an inference are judgments, not
propositions.  Nonsense.  The conclusion of a good inference follows
from the premises (and the inference rule), no matter what deontic
attitude observers may care to adopt.  If P is the conclusion of an
inference, it does not matter whether anybody asserts it - it remains
the conclusion either way.

Positive outcome: treat a:A as an inference schema.  This is closer to
actual practice: A is the conclusion.  We can treat a:A as a sequent.

Thinking in HoTT requires that we move from a denotational perspective
to an inferential one.

Under an inferential semantics, the notion of a model retains its
mathematical usefulness as a device for studying classes of
mathematical structures, but it loses its semantic role.  Models have
no explanatory significance; they are replaced by the notion of
inference as the primitive explanatory device.

Example: \&-intro.  Per ML, the conclusion is a ``judgment'':
(a,b):AxB.  Problem 1: illocutionary force is irrelevant; if there is
any judgment involve, it is the judgment that the conclusion follows
from the premises, not the judgment that the conclusion is true.  But
that kind of judgment we call inference, or maybe judgment as to the
validity of inference.  Either way, the conclusion does not involve
force.  Problem 2: informally we want to prove propositions like AxB,
not inferences like (a,b):AxB.  So we have two ways of reading
``conclusion''.  If we treat our calculus as a sequent calculus, then
the conclusion is the sequent (a,b):AxB.  If we treat it as a
propositional calculus (nat deduction), then the conclusion is AxB.

But type theoretic inference \emph{must} be treated as a sequent
calculus?

The four forms of judgment become inference schemata.

Definitional equality ``judgments'': \(a\equiv b:A\).  This is a
sequent, but its antecedent clause looks funny.  The key point to
remember is that \(\equiv\) (the ``coreference operator''), is like
``:'': both are operators of the metalanguage, not the language.  So
\(a\equiv b:A\) just means that \(a\) and \(b\) are co-referential
(classical perspective), or mutually substitutable (inferential
perspective): both \(a:A\) and \(b:B\) hold.  We really should not
call \(\equiv\) an equality operator, since we want to reserve that
term for the type-forming operator \(=\).  We could call it the
co-reference operator, but that smuggles in the notion of reference.
Howsabout ``substitution'' operator?  On the other hand, we can think
of \(a\equiv b\) as expressing an inference rule just like \(a:A\)
expresses an inference rule.  So maybe we should call it the
``substitution inference operator''.  Then \(a\equiv b:A\) comes out
as a complex sequent, one whose antecedent clause is a substitution
inference rule.  A special case: \(a\) and \(b\) are mutually
substitutable, and \(A\) is inferable either.

Truth is harmful insofar as it beguiles us into thinking that proof of
propositions involves a truth property.  At least for beginners, this
obscures the ``true'' nature of reasoning in type theory.

\section{Relation Types}

Assume type \(A\), i.e. read \(x=y\) as \(x=_Ay\), and
e.g. \(\prod\limits_{x,y}\) as \(\prod\limits_{x,y:A}\).

\begin{align}
  &\prod\limits_{x,y}\mathcal{P}^{\mathcal{U}} & \textit{function type of families}\ (A\to A\to\mathcal{U}) \\
  &\circ:\prod\limits_{x,y}\mathcal{P}^{\mathcal{U}} & \textit{arbitrary function (element) of family type} \\
  &a\circ b:\mathcal{P}^{\mathcal{U}} & a\circ b\ \textit{is a}\ \textsf{relation type} \\
  &a\circ a:\mathcal{P}^{\mathcal{U}} & a\circ b\ \textit{is a}\ \textsf{reflexive}\ \textit{relation type} \\
  &\diamond:\prod\limits_x (x\circ x) & \textit{constructor for reflexive relations} \\
  &\diamond x:(x\circ x) & \textit{canonical element of reflexive relation}\ \textsf{at}\ x \\
  &a:(x\circ x) & \textit{(arbitrary) element of reflexive relation}\ \textsf{at}\ x \\
  &\reflectbox{\lambda}(x\circ x) & \textit{(arbitrary) element of reflexive relation}\ \textsf{at}\ x 
\end{align}

\begin{align}
&\prod\limits_{p:x\circ x}\mathcal{P}^{\mathcal{U}} & \textit{family type built on reflexive relation} \\
&C\colon\!\prod\limits_{p:x\circ x}\mathcal{P}^{\mathcal{U}};\ C(\diamond x) & C(\diamond x)\ \textit{is a type} \\
&\diamond x\rvert\prod\limits_{p:x\circ x}\mathcal{P}^{\mathcal{U}} & \textit{same as prev}
\end{align}

\begin{align}
  &C\colon\prod\limits_{x,y}(x\circ y)\to\mathcal{P}^{\mathcal{U}} & \textit{2nd order family of families} \\
  &C\colon\prod\limits_{x,y}\prod\limits_{p:x\circ y}\mathcal{P}^{\mathcal{U}} & \textit{ditto} \\
  &C(a,b)\colon(a\circ b)\to\mathcal{P}^{\mathcal{U}} & \textit{family of implications over reln proofs} \\
  &C(a,b)\colon\prod\limits_{p:a\circ b}\mathcal{P}^{\mathcal{U}} & \textit{ditto} \\
  &C(a,a)\colon(a\circ a)\to\mathcal{P}^{\mathcal{U}} & \textit{family of implications over refl proofs} \\
  &C(a,a)\colon\prod\limits_{p:a\circ a}\mathcal{P}^{\mathcal{U}} & \textit{ditto} \\
  &\diamond a\colon a\circ a & \diamond a \ \textit{is a reflexivity proof} \\
  &C(a,a)(p):\mathcal{P}^{\mathcal{U}} & \textit{one type per proof}\ p\ \textit{of}\ a\circ a \\
\intertext{\(C(a,a,p)\) is an inference type expressing \textit{modus ponens}: if \(a\circ a\) then \(P\); but \(p\) proves \(a\circ a\); hence \(P\).}
  &C(a,a)(\diamond a):\mathcal{P}^{\mathcal{U}} & \textit{canonical case:}\ \diamond a\colon a\circ a \\
  &C(a,a,\diamond a):\mathcal{P}^{\mathcal{U}} & \textit{ditto} \\
&x:C(a,a,\diamond a) & \textit{arbitrary proof of}\ C(a,a,\diamond a) \\
\intertext{Logically, \(x\) validates the inference \(C(a,a,\diamond a)\).}
\end{align}

%% &\reflectbox{\lambda}A\to B & \textit{\reflectbox{\lambda}A\to B : A\to B} \\
%% &\reflectbox{\lambda}\mathbb{N} & \textit{an arbitrary natural number}

\section{Equality}

For any type \(A\) and \(x,y:A\), we can specify

\begin{align}
  &\prod\limits_{a:A}A
\end{align}

Interpreted computationally (functionally), this expresses
the type of functions that take an element of a type as input and
yield an element of the same type as output.

Interpreted logically, it expresses the type of implications that say:

\begin{quote}
\textit{for any proof of a proposition we can produce a proof of the proposition}
\end{quote}

On either interpretation, this type seems trivial; it looks like the
Identity type.  But that's not the right interpretation.  It's neither
the Identity function, nor the Identity implication; it does not
merely return the same element it is given, nor does it give the
premise as the conclusion of an implication.  Rather, as a function it
takes a particular element of \(A\) and yields \emph{some} (possibly
other) element of \(A\).  As an implication, it says precisely that
given a \emph{particular} proof of a proposition, we can find some
\emph{arbitrary} proof of the proposition - which may be a different
proof.  In other words under either interpretation it expresses a kind
of move from the particular to the general.

\textbf{FIXME:} what should we call this kind of type?  It's neither
an Identity type nor an Equality type, but is related to both.  A
``proof equality type''?

Now consider:

\begin{align}
  &\prod\limits_{a:A}A\to\mathcal{U} & \textit{read:}\ \prod\limits_{a:A}(A\to\mathcal{U})
\end{align}

This type expresses the notion of a \emph{family of types}.  It is the
type of such type families; an element of this type is a ``type
family'', which is just a function that produces types.  Any given
element of this type, such as
\(C\colon\prod\limits_{a:A}A\to\mathcal{U}\), will take an element
(proof) of a type, and yield (imply) not another element but another
type - one that is dependent of the element used to produce it.  So if we
have \(b\colon A\), then \(C(b)\colon\mathcal{U}\), meaning \(C(b)\)
is a type dependent on \(b\).

\begin{quote}
Note that we can treat \(C(b)\) as an inference type expressing modus
ponens: \(A\to\mathcal{P}^{\mathcal{U}}\); but \(a:A\); therefore \(\mathcal{P}^{\mathcal{U}}\).
\end{quote}

This simple type pattern is surprisingly important in the HoTT
presentation of equality.  Compare:

\begin{align}
  &\prod\limits_{x,y:A}\mathcal{P}^{\mathcal{U}} & \textrm{same as}\ A\to A\to\mathcal{U} \label{pi:id} \\
  &\prod\limits_{x,y:A}(x=_Ay) &\textit{fns returning a proof for a pair} \label{pi:idp} \\
  &\prod\limits_{x,y:A}(x=_Ay)\to\mathcal{U} & \textit{fns returning a fn on proofs for a pair} \label{pi:idppfam} \\
  &\prod\limits_{p:x=_Ay}(x=_Ay) \label{pi:idpp} & \textit{fns returning a proof for a proof} \\
  &\prod\limits_{p:x=_Ay}(x=_Ay)\to\mathcal{U} & \textit{fns returning a fn on proofs for a proof} \label{pi:idfam}
\end{align}

Here (\ref{pi:id}) is the type of the family of Equality (Identity)
types in HoTT, which sets \(Id:A\to A\to\mathcal{U}\).



\section{Inferences as Types}
Type theory subsumes classical (traditional?) logic in its entirety -
not just propositions as types, but terms, propositions, and
inferences as types.  This also means that the truth of a proposition
is equivalent to the validity of an inference.

\subsection{Examples}
Assume \(\mathcal{P:U}\).
\medskip

\noindent Extensional glosses:

\begin{flalign}
&(x=_Ay) & x=y \\
&(x=_Ay)\to\mathcal{P} & \textit{if}\ x=y \ \textit{then}\ \mathcal{P}\ \textit{is a proposition} \\
&\prod\limits_{x,y:A} (x =_A y) \to \mathcal{P} & \forall x,y, \textit{if x=y then}\ \mathcal{P}\ \textit{is a proposition} \label{eq:univ} \\
&C:\prod\limits_{x,y:A} (x =_A y) \to \mathcal{P}  & C\ \textit{proves proposition 3} \\
&C(a,b) & \textit{if } (a=_Ab)\ \textit{then }\ \mathcal{P}\ \textit{is a proposition} \label{eq:inst} \\
&b,a\raisebox{-4pt}{\(\biggr\rvert\)}\prod\limits_{x,y:A} (x =_A y) \to \mathcal{P} & \textit{same as proposition 5} \\
&p:(a=_Ab) & \textit{p proves}\ a=_Ab \\
&p\big\rvert(a=_Ab)\to\mathcal{P} & (a=_Ab)\to\mathcal{P};\ \textit{but}\ p;\ \textit{therefore}\ \mathcal{P} \label{eq:pab}\\
&C(a,b)(p) & \mathcal{P}\ \textit{is a proposition}\label{eq:cabp} \\
&p,b,a\raisebox{-4pt}{\(\biggr\rvert\)}\prod\limits_{x,y:A} (x =_A y) \to \mathcal{P} & \textit{same as (\ref{eq:cabp})}\label{eq:inference}
\end{flalign}

\noindent Intensional gloss: equation (\ref{eq:pab}) may be glossed as follows:

\begin{quote}
if \(a=b\) is proved, then \(\mathcal{P}\); but \(p\) proves \(a=b\), therefore \(\mathcal{P}\).
\end{quote}

\noindent (\ref{eq:inference}) can be similarly glossed, but to make it fully and hideously explicit, we would have to add clauses expressing the inference from the universal closure expressed by (\ref{eq:univ}) and its instantiation expressed by (\ref{eq:inst}).  That tedious exercise is left to the reader.



%% \begin{flalign}
%% &(x=_Ay) & x=y \\
%% &(x=_Ay)\to\mathcal{P} & \textit{if}\ x=y \ \textit{then}\ \mathcal{P}\ \textit{is a proposition} \\
%% &\prod\limits_{x,y:A} (x =_A y) \to \mathcal{P} & \forall x,y, \textit{if x=y then}\ \mathcal{P}\ \textit{is a proposition} \\
%% &C:\prod\limits_{x,y:A} (x =_A y) \to \mathcal{P}  & C\ \textit{proves proposition 3} \\
%% &C(a,b) & \textit{if } (a=_Ab)\ \textit{then }\ \mathcal{P}\ \textit{is a proposition} \\
%% &b,a|\prod\limits_{x,y:A} (x =_A y) \to \mathcal{P} & \textit{same as proposition 5} \\
%% &p:(a=_Ab) & \textit{p proves}\ a=_Ab \\
%% &C(a,b)(p) & \mathcal{P}\ \textit{is a proposition} \\
%% &p,b,a|\prod\limits_{x,y:A} (x =_A y) \to \mathcal{P} & \textit{same as proposition 8}
%% \end{flalign}


Note: 8 and 10 use an unorthodox notation that I'm calling the
"tokenization operator".  The idea is that for any type A, "|" picks
out an element of the type, and we then use prefix notation to
indicate application. (We might also split out a tokenization
operator, e.g. Λ and define a "pipeline" application operator.)  This
allows us to dispense with (some) intermediate variables, just like
the lambda operator allows us to dispense with function names.  In
8, \(|(a=b)\to\mathcal{P}\) picks out an element of the function type,
and prefixing p to it indicates application of that element to p.  So
8 is a (nearly) variable-free version of:

    \[C:(a=b)\to\mathcal{P}; C(p)\]

Note that 10 could be written more explicitly:

\[a,b:A, p:(a=b), p\rvert b,a\raisebox{-4pt}{\(\biggr\rvert\)}\prod\limits_{x,y:A}(x=Ay)\to\mathcal{P}\]


which should be read "outward": feed a then b then a proof of a=b to the \(\Pi\) function.

The glosses for 8-10 here ignore information in the formulae; the gloss only applies to the "result" of the type formation operation.  They can also be glossed intensionally as instances of modus ponens
(\(A\to B; but A, therefore B\)):

\begin{quote}
    If a=b is proven, then P; but p proves a=b, therefore P.
\end{quote}

Here I think the advantage of having a tokenization operator like "|" is evident; it makes it easier to read the intensional sense of 8 and 10 directly off the formula.  Well, not entirely; we still use vars a, b, and p.   They could be eliminated using the tokenization operator with a \(\Sigma\) type, but that's a topic for another post.

What this is intended to show is that Curry-Howard is not limited to propositions.  Types may encode terms, propositions, or inferences.   Those that express inferences may also be construed as expressing propositions (namely, the concluding proposition of the inference); those that express propositions may be construed as expressing terms.  The latter corresponds to the fact that embedded propositions function as terms: in "if P then Q", both P and Q are propositional variables but they act as terms within the conditional.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Z for set theory}

Example from section 1.6 Dependent Pair types (Σ-types), p. 30:

\begin{quote}
``The way to construct elements of a dependent pair type is by
pairing: we have \((a,b) : ∑(x:A) B(x)\) given \(a : A\) and \(b :
B(a)\).''
\end{quote}

In Z: \[ {(a,b) | a\in A \land b\in P(a)} \]

The critical point being that Z makes the ``such that'' explicit.

ML, ``An intuitionistic theory of types'' p. 4:
\begin{quote}
``A third function of \((\sum x\in A)B(x)\) is to represent `the type
of all objects \(a\) of type \(A\) \emph{such that} \(B(a)\)...''
\end{quote}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Meaning Explanation}

\begin{quote}
``What we do here is meant to be closer to ordinary mathematical practice. We will avoid keeping form and meaning (content) apart. Instead we will at the same time display certain forms of judgement and inference that are used in mathematical proofs and explain them semantically.'' ML, Intuitionistic Type Theory p. 2
\end{quote}

Note: explain them \emph{semantically}; this is ambiguous.  Does it
mean explain them in representational terms, or does it mean give a
meaning explanation?

TODO:  Dummett quote


\end{document}
