\documentclass[12pt,toc]{tufte-handout}
%% \documentclass[reqno,12pt]{tufte-book}
%% \usepackage{trace}
%% \documentclass[reqno,12pt]{article}

\usepackage{draftwatermark}

% BLACK & WHITE
\input{opt-black-white}

% FORMATTING DEPENDENT ON PAPER SIZE
\input{opt-letter}

\usepackage{etex}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% packages included by original hott main.tex

%%% For table {tab:theorems}
\usepackage{pifont}

%%% Multi-Columns for long lists of names
\usepackage{multicol}

\usepackage{graphicx}
\usepackage{comment}

\usepackage{fancyhdr} % To set headers and footers

%% \usepackage{nextpage} % So we can jump to odd-numbered pages

\usepackage{amssymb,amsmath,stmaryrd,mathrsfs,wasysym}
\usepackage{enumitem,mathtools,xspace}
%% \numberwithin{equation}{subsection}

\usepackage{xstring} % For generating singluars and plurals in \backref

%% \usepackage{xcolor,mdframed}
\usepackage{xcolor} % For colored cells in tables we need \cellcolor
\usepackage{wallpaper} % For the background image on the cover page

\usepackage{booktabs} % For nice tables
\usepackage{array} % For nice tables

\definecolor{linkcolor}{rgb}{\OPTlinkcolor}
\usepackage{aliascnt}
\usepackage[capitalize]{cleveref}
\usepackage[all,2cell,cmtip]{xy}
\UseAllTwocells
%\usepackage{natbib}
\usepackage{braket} % used for \setof{ ... } macro

\usepackage{tikz}
\usetikzlibrary{decorations.pathmorphing,arrows}

\usepackage{etoolbox}           % hacking commands for TOC

%% \usepackage{mathpartir}         % for formal.tex appendix, section 3

\usepackage[numbered]{bookmark} % add chapter/section numbers to the toc in the pdf metadata
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{framed}
\usepackage[standard,framed]{ntheorem}
%% \newtheorem{theorem}{Theorem}
%\newtheorem{cor}{Corollary}
%\newtheorem{lem}{Lemma}
%% \newtheorem*{defn}{Definition}
%% \theoremstyle{remark}
%% \newtheorem{remark}{Remark}
%% \newtheorem*{commentary}{Commentary}

%% \theoremclass{Remark}
%% \theoremstyle{break}
%% \newtheorem{note}{Note}[section]

\theoremstyle{plain}
\theorembodyfont{\upshape}
\theoremsymbol{\ensuremath{\ast}}
\theoremseparator{}
%% \newtheorem{ednote}{Ed. note}[section]
\newframedtheorem{ednote}{Ed. note}[section]

\newtheorem*{todo}{TODO}
%% \newtheorem{eg}{Example}

\input{macros}

\usepackage{appendix}

%% \usepackage{csquotes}

\usepackage{setspace}

%% broken (doesn't work with tufte-handout):
\usepackage{zed-csp}
%% broken:
%% \usepackage{ltcadiz-fam}

\usepackage{fontspec}
%% \usepackage{xltxtra,xunicode}
\defaultfontfeatures{Scale=MatchLowercase}

%% \defaultfontfeatures{Scale=MatchLowercase}
%% \setmainfont[Mapping=tex-text]{Times New Roman}
%% \setsansfont[Mapping=tex-text]{Arial}
%% \setmonofont{Courier}

\setmainfont[Ligatures=TeX]{TeX Gyre Bonum}
\setromanfont[Ligatures=TeX]{TeX Gyre Bonum}
\setsansfont[Ligatures=TeX]{TeX Gyre Adventor}
\setmonofont[Ligatures=TeX]{TeX Gyre Cursor}


%% \setmainfont[Mapping=tex-text]{Minion Pro}
%% \setromanfont[Mapping=tex-text]{Minion Pro}
%% \setsansfont[Mapping=tex-text]{TeX Gyre Heros}

%% Bugfix: see https://code.google.com/p/tufte-latex/issues/detail?id=64
% Set up the spacing using fontspec features
\renewcommand\allcapsspacing[1]{{\addfontfeature{LetterSpace=15}#1}}
\renewcommand\smallcapsspacing[1]{{\addfontfeature{LetterSpace=0.0}#1}}

\usepackage{epigraph}
\setlength{\epigraphwidth}{.8\textwidth}

%% general symbols - degree, etc.
%% \usepackage{gensymb}

\usepackage [english]{babel}
\usepackage [autostyle, english = american]{csquotes}

%% nice double-stroke fonts
\usepackage{dsfont}

% Small sections of multiple columns
\usepackage{multicol}

% Provides paragraphs of dummy text
\usepackage{lipsum}

% The units package provides nice, non-stacked fractions and better spacing
% for units.
\usepackage{units}

%\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ...

\usepackage{xfrac}

\usepackage{hyperref}
\hypersetup{
  bookmarks=true,         % show bookmarks bar?
  bookmarksdepth=3,
  unicode=true,          % non-Latin characters in Acrobat’s bookmarks
  pdftoolbar=true,        % show Acrobat’s toolbar?
  pdfmenubar=true,        % show Acrobat’s menu?
  pdffitwindow=false,     % window fit to page when opened
  pdfstartview={FitH},    % fits the width of the page to the window
  pdftitle={Intuition and Exponentiation},    % title
  pdfauthor={G. A. Reynolds},     % author
  pdfsubject={Mathematics},   % subject of the document
  pdfcreator={G. A. Reynolds},   % creator of the document
  pdfproducer={Producer}, % producer of the document
  pdfkeywords={Exponentiation} {Mathematics}
  pdfnewwindow=true,      % links in new window
  colorlinks=true,       % false: boxed links; true: colored links
  linkcolor=blue,          % color of internal links
  citecolor=blue,        % color of links to bibliography
  filecolor=magenta,      % color of file links
  urlcolor=cyan           % color of external links
}

%% \usepackage[
%% bibstyle=numeric,
%% citestyle=authoryear,
%% hyperref,
%% bibencoding=utf8,
%% backref=true,
%% backend=biber]{biblatex}

%% http://tex.stackexchange.com/questions/66778/citation-alias-with-multibib-and-natbib
%% \makeatletter
%% \def\@mb@citenamelist{cite,citep,citet,citealp,citealt,citepalias,citetalias}
%% \makeatother

%% http://stackoverflow.com/questions/2496599/how-do-i-cite-the-title-of-an-article-in-latex
\defcitealias{z-iso-13568}{ISO 13568:2002 Information technology -- Z formal specification notation --
  Syntax, type system and semantics}

\usepackage{tikz}
\usepackage[markings,customcolors]{hf-tikz}
\usetikzlibrary{%
  arrows%
  ,calc%
  ,decorations.text%
  ,decorations.pathreplacing%
  ,fadings%
  ,positioning
  ,shapes.geometric%
}

\usepackage{tikz-3dplot}

\usepackage{pgfplots}
\pgfplotsset{height=7cm,compat=1.9}

\usepackage{tkz-euclide}
\usetkzobj{all}

%% prettier integral syms, but broken on miktex
%% \usepackage{esint}


%% \usepackage{MnSymbol}
%% \usepackage[misc]{ifsym}

%% \usepackage{morefloats}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{HoTT Types}
%% \\
%% \Large Derived from the HoTT Book}
\author{}
%\date{}                                           % Activate to display a given date or no date

%%%%%%%%%%%%%%%%
%% tufte-latex customizations

\makeatletter
\let\runauthor\@author
\let\runtitle\@title
\makeatother

%% running headers
\newcommand{\changefont}{%
  \fontsize{7}{9.5}\selectfont
}
\fancypagestyle{plain}{
  \fancyhead[LO,LE]{\leftmark }
  \fancyhead[RO,RE]{\rightmark}
  \fancyfoot[CO,CE]{\thepage}
  \fancyfoot[LE]{\textsc{\runtitle}}
  \fancyfoot[RO]{\textsc{\runtitle}}
  \renewcommand{\headrulewidth}{0pt}
  \renewcommand{\footrulewidth}{0pt}
}
\pagestyle{plain}

\def\chpcolor{blue!45}
\def\chpcolortxt{blue!60}
\def\sectionfont{\LARGE}

\setcounter{secnumdepth}{5}
\setcounter{tocdepth}{5}        % sections and subsections for the toc

\makeatletter
%% Section:
\def\@sectionstrut{\vrule\@width\z@\@height12.5\p@}
\def\@makesectionhead#1{%
  {%\par\vspace{20pt}%
    \parindent -10pt\raggedleft\sectionfont
    %% \colorbox{\chpcolor}{%
    %%   \parbox[t]{90pt}{\color{white}\@sectionstrut\@depth4.5\p@\hfill
    %%     \ifnum\c@secnumdepth>\z@\thesection\fi}%
    %% }%
    \vspace{10pt}%
    \begin{minipage}[t]{\textwidth}%{\dimexpr\textwidth-90pt-2\fboxsep\relax}
      \@sectionstrut\hspace{-15pt}\textit{\textbf\Huge #1}
    \end{minipage}\par
    \vspace{5pt}%
  }
}
%% \def\@makesectionhead#1{%
%%   {\par\vspace{20pt}%
%%    \parindent 0pt\raggedleft\sectionfont
%%    \colorbox{\chpcolor}{%
%%      \parbox[t]{90pt}{\color{white}\@sectionstrut\@depth4.5\p@\hfill
%%        \ifnum\c@secnumdepth>\z@\thesection\fi}%
%%    }%
%%    \begin{minipage}[t]{\dimexpr\textwidth-90pt-2\fboxsep\relax}
%%    \color{\chpcolortxt}\@sectionstrut\hspace{5pt}\textbf{#1}
%%    \end{minipage}\par
%%    \vspace{10pt}%
%%   }
%% }
\def\section{\@afterindentfalse\secdef\@section\@ssection}
\def\@section[#1]#2{%
  \ifnum\c@secnumdepth>\m@ne
  \refstepcounter{section}%
  \addcontentsline{toc}{section}{\protect\numberline{\thesection}#1}%
  \else
  \phantomsection
  \addcontentsline{toc}{section}{#1}%
  \fi
  \sectionmark{#1}%
  \if@twocolumn
  \@topnewpage[\@makesectionhead{#2}]%
  \else
  \@makesectionhead{#2}\@afterheading
  \fi
}
\def\@ssection#1{%
  \if@twocolumn
  \@topnewpage[\@makesectionhead{#1}]%
  \else
  \@makesectionhead{#1}\@afterheading
  \fi
}
\makeatother

%%%%%%%%%%%%%%%%
%% macros

\newenvironment{important}[1][]{%
  \begin{mdframed}[%
      backgroundcolor={red!15}, hidealllines=true,
      skipabove=0.7\baselineskip, skipbelow=0.7\baselineskip,
      splitbottomskip=2pt, splittopskip=4pt, #1]%
    \makebox[0pt]{% ignore the withd of !
      \smash{% ignor the height of !
        \fontsize{32pt}{32pt}\selectfont% make the ! bigger
        \hspace*{-19pt}% move ! to the left
        \raisebox{-2pt}{% move ! up a little
          {\color{red!70!black}\sffamily\bfseries !}% type the bold red !
        }%
      }%
    }%
}{\end{mdframed}}

%% reversed integral sign
\makeatletter
\providecommand*{\curv}{%
  \mathrel{%
    \mathpalette\@curv\int
  }%
}
\newcommand*{\@curv}[2]{%
  \reflectbox{$\m@th#1#2$}%
}
\makeatother

%% \def\LaTeX{%
%%   L\kern-.36em
%%   {\setbox0=\hbox{T}%
%%     \vbox to \ht0{\hbox{\the\scriptfont0 A}\vss}}%
%%   \kern-.15em
%%   \TeX
%% }

%%%%%%%%%%%%%%%%

\newcommand\cspace{coordinate space}
\newcommand\Cspace{Coordinate space}
\newcommand\CSpace{Coordinate Space}

\newcommand\dspace{design space}
\newcommand\Dspace{Design space}
\newcommand\DSpace{Design Space}

\newcommand\Omg{\(\Omega\)}
\newcommand\sccs{standard cartesian coordinate space}
\newcommand\origin{\((0,0)\)}
\newcommand\ab{\((a,b)\)}

\newcommand\atypeA{\ensuremath{(a : A)}}

%% \newcommand\N{\(\mathds{N}\)}
%% \newcommand\R{\(\mathds{R}\)}
%% \newcommand\RR{\(\mathds{R}\times\mathds{R}\)}
%% \newcommand\Rtwo{\(\mathds{R}^2\)}
%% \newcommand\Z{\(\mathds{Z}\)}


\def\HoTT{%
  H\kern-.7pt
  {\tiny\raisebox{1pt}{o}}%
  %% {\setbox0=\hbox{T}%
  %%  \vbox to \ht0{\vss\hbox{\the\scriptfont0 o}\vss}}%
  \kern-1.5pt
  TT}

\def\HoTTB{%
  the H\kern-.7pt
  {\tiny\raisebox{1pt}{o}}%
  %% {\setbox0=\hbox{T}%
  %%  \vbox to \ht0{\vss\hbox{\the\scriptfont0 o}\vss}}%
  \kern-1.5pt
  TT Book
}

\newcommand\ML{Martin-L\"{o}f}

\newcommand\TTh{Type Theory}

\newcommand\tth{type theory}

\includeonly{%
introduction%
,preliminaries
,niceties
,pragmatism
,foundations
,types
,curry-howard
,proof
,semantics
,math
,brandom
,hotttypes
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%% \ifx\traceon\undefined \tracingall \else \traceon \fi

\maketitle

\begin{abstract}
  Currently this doc contains a (mildly organized) set of notes
  followed by the intro and chapter 1 from the
  \href{http://homotopytypetheory.org/book/}{HoTT Book}.  Eventually
  (maybe) the intro and chapter 1 will contain annotations, comments,
  additional examples, etc., but I have not started that yet, so if
  you are already familiar with the text you need not read them -- I
  haven't (so far as I recall) changed anything.

  The idea is to winnow out some of the strictly mathematical stuff
  leaving the core ``philosophical'' stuff, and annotate the text with
  some comments and quotes from Martin-L\"{o}f, Brandom, etc.  Or
  maybe leave the math stuff in, but annotate it with more detailed
  explanation and examples in programming languages.  In any case the
  purpose is to more fully articulate the link between HoTT's ideas of
  type and judgment (etc.) to the philosophical debates about
  language, assertion, proposition from which they emerged.  Why?
  Because I find those bits of the HoTT a little murky, and
  philosophers like Brandom have a lot to say about the issues.  Also,
  to show more clearly how type theory differs from set theory and
  classic logic.  Another goal is to provide more practical guidance
  to programmers interested in exploring dependent types.
\end{abstract}

\tableofcontents
%% \setcounter{tocdepth}{2}        % chapters, sections, and subsections for the
%%                                 % metadata of the pdf
%% \cleartooddpage[\thispagestyle{empty}]

%% \mainmatter % Turn on roman page numbers and numbered chapters

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\include{pragmatism}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Logics}
\label{sect:logics}

\begin{description}
\item [Traditional] terms are primitive; propositions are combinations of terms; judgments apply to proposotions
\item [Modern: classic] LEM, AC, etc.
\item [Modern: intuitionistic]
\item [Expressivism]  Brandom's version: propositions are primitive; relation to inferential semantics; Price's global expressivism
\end{description}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\include{proof}

\include{semantics}

\include{math}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\include{foundations}

\include{types}

\include{curry-howard}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Assertion and Judgment}
\label{sect:assertionjudgment}

The account of judgment offered in the HoTT Book doesn't really work.
Ditto for Martin-L\"{o}f's account.  For example, it makes sense to
say ``P is a proposition'', but it doesn't make sense to say ``P is a
judgment''.  That's because judgment is a act, something one does.

On the other hand, ``judgment'', like ``proposition'', can be treated
as a verbal noun or as a ``plain'' noun.  Saying ``P is a
proposition'' is usually taken to mean that P refers to what has been
proposed.  There is no obvious reason not to treat ``P is a judgment''
in a similar manner: P refers to what has been judged.

However, there is a difference.  Judging a proposition (what was
proposed) amounts to \textit{evaluating} what was proposed, as good or
bad, true or false, or whatever.  By contrast, proposing a proposition
amounts to merely exhibiting it for consideration.  This arguably
involves an implicit evaluation - to propose a proposition is to
implicitly claim that it is good, or true, etc.  But proposing does
not involve offering an evaluation that is distinct from what is
proposed, whereas judgment does.  The two are distinct kinds of speech
act, and refering to the content of a speech act is not the same as
referring to the speech act itself.

Furthermore, it is not correct to treat the nominal sense of
``judgment'' as being the content, what has been judged.  The nominal
sense of ``judgment'' refers to the act of judgment itself, and not
the proposition judged.

Actually, by the same reasoning it is not correct to say that the
nominal sense of ``proposition'' is what-is-proposed; rather, it is
the act proposing, nominalized.  This makes perfect sense when you
consider that ``proposing'' can also be nominalized; ``the proposing''
is another way of saying ``the proposition''.

The same goes for all -tion words: suggestion, opposition, etc.  In
each case, the word can refer to the doing, or to what is done, and
what is done is always the act of doing itself -- not the subject or
object of the doing.

This suggests we should make a distinction between, for example, the
content of a proposition and ``proposition''.  But this term seems to
be a special case; it has the usual plain noun sense of
what-was-proposed, the usual verbal sense of ``proposing'', but also
the nominalized verbal sense of ``act of proposing''.

(But then the same considerations apply to ``judgment''.  The
difference must go back to semantics.)

\begin{remark}
  The Arabic grammatical tradition captures this distinction
  beautifully, mainly because the structure of the language makes it
  simple to do so.
\end{remark}

Or put it this way: when we judge a proposition like ``2+2=4'' to be
true, the what-was-judged is not ``2+2=4'' but the truth of ``2+2=4''.

\begin{remark}
  But how is this different from ordinary predication, like ``The
  triangle is red'' as a proposition?  Should we say that what is
  proposed is not that the triangle is red, but the redness of the
  triangle?  No, since we're treating it as a propostion, and the
  whole thing is proposed (exhibited).  If we judge it to be true,
  then again the judgment 
\end{remark}

So saying ``P is a judgment'' is incoherent if P is taken to refer to
nothing more than what is proposed.  If P refers to a claim of the
form ``X is true'' (or good, etc.), then ``P is a judgment'' seems to
make more sense; but it doesn't, really.  P still refers to an
unasserted content; to make sense, we would have to say something like
``P is a judgment when asserted''.  More explicitly, ``'X is true' is
a judgment'' (or better, ``'X is true' expresses a judgment'') only
\textit{exhibits} ``X is true'', which is a proposition, not a
judgment.  As a proposition it expresses a judgment; but when embedded
(equivalently, quoted) it does not express anything.

\begin{remark}
  Compare: ``Snow is white'' iff snow is white.  The quoted bit is a
  name of the sentence; it counts as a \textit{mention} of the
  sentence, which has no force.  The unquoted version of same is the
  sentence itself; it counts as a \textit{use} of the sentence, which
  has assertional force.  Obviously, the occurances of ``P'' in ``P is
  a proposition'' and ``P is a judgment are names of a proposition and
  thus mentions.  So they have no force.
\end{remark}

The key point is Frege's point: the content of a proposition is
distinct from the force of the utterance.  That means that P in ``P is
a proposition'' is unasserted, just as it is when embedded, as in ``If
P then Q''.  The truth of ``P is a proposition'' is independent of the
truth of P.

So even if we take the act of declaring ``P'' to be an act of
judgment, it does not follow that a reference to P is a reference to
the act of judging that P.  Hence there is no way to make ``P is a
judgment'' work.  If we take P to refer to what was judged, that again
is a proposition (or propositional content), so ``P is a judgment'' is
incoherent.

\begin{remark}
  We can assert that P, and we can assert P.  We can judge that P, but
  we cannot judge P.  I don't think this is a mere grammatical
  distintion; I think it reflects a genuine semantic difference.
\end{remark}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{What's the Big Deal about Equality?}
\label{sect:equality}

\begin{ednote}
  Equality is arguably the most important concept of \HoTT{}, as far
  as I can tell, because of the ``Univalence Axiom''.
\end{ednote}

``In the intensional version of the theory, with which we are
concerned here, one thus has two different notions of equality:
propositional equality is the notion represented by the identity
types, in that two terms are propositionally equal just if their
identity type IdA(a,b) is inhabited by a term. By contrast,
definitional equality is a primitive relation on terms and is not
represented by a type; it behaves much like equality between terms in
the simply-typed lambda-calculus, or any conventional equational
theory.

If the terms a and b are definitially equal, then (since they can be
freely substituted for each other) they are also propositionally
equal; but the converse is generally not true in the intensional
version of the theory''\cite{awodey_tth}

``The constructive character, computational tractability, and proof-
theoretic clarity of the type theory are owed in part to this rather
subtle treatment of equality between terms, which itself is
expressible within the theory using the identity types IdA(a, b).''\cite{awodey_tth}

%%%%%%%%
\subsection{Substitution}
\label{subs:substitution}

As the quote from Awodey above suggests, the concept of
substitutability plays a basic role.

\begin{ednote}
  Compare substitution in lambda calculus, and in Brandom's model.
  Maybe something about combinatory logic and the elimination of
  variables?
\end{ednote}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Expressivity}
\label{sect:expressivity}

Instead of ``P is a proposition'' etc. we should say ``P expresses a proposition''.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Determinism}
\label{sect:determinism}

Hypothesis: classical math with LEM and AC is inherently
non-deterministic.  Constructive math(s) and logic(s) that discard LEM
and AC are deterministic.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Modality}
\label{sect:modality}

Classic proofs (that use LEM) are modal.  Consider the way a classic
LEM-dependent proof works.  You start by stating the hypothesis: P is
true.  You assume that P is not true; then you derive a contradiction.
The conclusion is not merely that P is true, however; it is that P
\textit{must} be true.

Constructive proofs, by contrast, are not modal.  They do not say what
must be the case, they say what \textit{is} the case.  Or rather, they
\textit{show} what is the case.  (I leave aside the question of
whether what is, is necessary.)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Habeus Corpus Logics}
\label{sect:habeus}

The principle of ``Habeus Corpus'', from the Latin ``(You shall) have
the body'', was once enshrined as a fundamental principle of
Anglo-American law.  It was used to force the State to present a
detainee in person before the court, back in the days when we
occasionally had the temerity to question the wisdom of the State when
it tried to disappear people.

Type theory, and constructive logics generally, operate under a writ
of habeus corpus that is permanently in effect.  Except that this writ
requires the production not of a detainee, but of a witness.  If you
claim to have a proof, you must produce a witness who is competent to
testify to that fact.  So its actually more like a law of evidence,
but I'm too lazy to come up with a clever legalism to express that
idea.




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Frege}
\label{sect:frege}

{\todo The Frege Point; force v. content, etc.}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Martin-L\"{o}f}
\label{sect:ml}

{\todo Summarize ML's remarks on assertion, proposition, etc.}

\include{brandom}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{From Truth to Testimony}
\label{sect:truth}

We have propositions as types, and we have non-propositional types
like $\nat$.  There is an obvious conflict of intuitions here.
Propositions, like \(1>0\), have truth conditions; names like \(\nat\)
do not.  How can they be the same kind of thing?

I think the way out of this embarrassment is recognition that the
classic concept of truth is not relevant to type theory; or, in a more
positive vein, that only a deflationary or minimalist notion of truth
should be used in type theory.  In type theory one does not say that a
proposition \textit{is} true or false; instead one says that a
(propositional) type is proven or disproven, or that either it or its
negation has a witness\sidenote{Or, a ``maker'' or ``constructor''.}.  Instead of a concept of truth we have a
concept of testimony.  Of course, ordinarily witnesses testify as to
the truth of some proposition; but the witnesses of type theory do
more than that---or rather they do something else, namely they produce
or ``make'' the proposition.

%%%%%%%%
\subsection{Proof, Witness, Constructor}
\label{subs:pwc}

Type theory seems to have settled on an idiom; one says, for example,
that types have or do not have proofs or witnesses.  But there are
problems with both of these terms.  The former covers too much ground
since it includes non-constructive proofs.  The latter invokes a
misleading metaphor, since a witness testifies to the truth, whereas a
type-theoretic witness to a type constructs (makes, produces, etc.)
something.  In the case of propositional types, constructors ``make''
the proposition (in the sense that they are inferences that terminate
in the proposition); in the case of non-propositional types,
constructors make ``elements'' of the type, which serve as proxies(?) for
the type.

\begin{remark}
  Problem: here again propositional and non-propositional types behave
  differently.  Every proof of a proposition has the proposition as
  its conclusion; they are all ``the same'' because they all have that
  element in the same structural position.  But the proof of
  e.g. \(\nat\) is different.  For example, \(2\) is a witness for
  \(\nat\).  Or rather, anything that constructs \(2\) is such a
  witness.  What all such constructions have in common is \(2\), not
  \(\nat\).  So they are all clearly proofs of \(2\), but we want them
  to be proofs of \(\nat\).  How do we get there?
\end{remark}

%%%%%%%%%%%%%%%%
\include{niceties}

\section{The Language of \HoTT}
\label{sect:hottlang}

One way to think about mathematics and logic is in terms of objects,
structures, relations, and the like.  etc.

But one can also think of it in terms of vocabularies (or idioms,
etc.).  Then mastering a discipline is not just a matter grasping some
content, but also of acquiring practical mastery over a vocabulary.

The vocabulary of set theory has dominated mathematical discourse for
most of the last 100 years or so.  Starting in the late 1940s, a
competing vocabulary based on category theory began to emerge.  Today
it is not uncommon to see both vocabularies deployed in the same
discourse (lecture, paper).

{\todo Type theory as a vocabulary - mostly confined to logic, then
  computer science.  Etc.  \HoTT{} as the latest distinctive vocab. -
  covering both math and compsci, also regions of logic.
  Significantly different that both set theory and classic logic.}

\include{hotttypes}

\clearpage
\appendix
\begin{appendices}

\include{introduction}

\include{preliminaries}

%%%% Bibliography
%% \bibliographystyle{halpha}
%% \phantomsection % black magic to get TOC to point to correct page
%% \addcontentsline{toc}{part}{\bibname}
%% \markboth{}{\textsc{Bibliography}}
%% {\renewcommand{\markboth}[2]{} % Prevent bibliography from resetting the header to something silly
%% \OPTbibliographyfont
\bibliography{references}
\bibliographystyle{plainnat}

\end{appendices}

\end{document}
